### 2020-06-04 writeup/transfer.tex
sec 3

	With high probability, the variance of the multi-task estimator $\hat{\beta}_t^{\MTL}$ equals the following

@@

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Recall that each task has sample size $n_i = \rho_i \cdot p$, for $i = 1, 2$.
	And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	We assume that for the target task, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma_i^2$, for $i=1,2$.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

In the isotropic model, we show that as we increase the distance between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL.
%We measure model dissimilarity as $\norm{\beta_1 - \beta_2}^2$, which is the distance between source and target in the isotropic model.
%Figure \ref{fig_model_shift} provides a simulation when $p = 200$.
%{The rest of parameter settings can be found in Appendix \ref{app_synthetic}.}
Our result below will provide an explanation to this phenomenon.
%Based on Theorem \ref{thm_model_shift}, we derive the transition threshold in the following proposition.
We introduce the following notations.
{\small\begin{align*}
	\Psi(\beta_1, \beta_2) = {\ex{\bignorm{\beta_1 - \beta_2}^2}} / {\sigma^2},  \quad \Phi(\rho_1, \rho_2) = \frac{(\rho_1 + \rho_2 - 1)^2}{\rho_1 (\rho_1 + \rho_2) (\rho_2 - 1)}.
\end{align*}}

@@

\begin{example}\label{ex_complement}
	We measure covariate shift by the matrix $M = \Sigma_1^{1/2} \Sigma_2^{-1/2}$.
	Assume that $\Psi(\beta_1, \beta_2) = 0$ for simplicity.
	We compare two cases: (i) when $M = \id$; (ii) when $M$ has $p/2$ singular values that are equal to $\lambda$ and $p/2$ singular values that are equal to $1 / \lambda$.
	Hence, $\lambda$ measures the severity of the covariate shift.
	Figure \ref{fig_covariate} shows a simulation of this setting by varying $\lambda$.
	We observe that as source/target sample ratio increases, the performance gap between the two cases increases.
\end{example}

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
To compare different choices of $M$ regarding the performance of $\hat{\beta}_t^{\MTL}$, let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M$ in decreasing order.
Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.
Consider the following bounded set
{\small\begin{align*}
		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_p\le \lambda_1 \le \mu_{\max}\right.},
\end{align*}}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.

\begin{proposition}[Covariate shift]\label{prop_covariate}
	Assume that $\Psi(\beta_1, \beta_2) = 0$ and $\rho_1, \rho_2>1$.
	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when the covariance shift matrix is equal to $M\in\cS_{\mu}$.
	We have that
	{\small\[ g(\mu\id) \le \bigbrace{1+ \bigo{\frac{\rho_2}{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]}
\end{proposition}
This proposition shows that when the source sample size is much larger than the target sample size, $\te(\hat{\beta}^{\MTL})$ is minimized when there is no covariate shift between source and target tasks.
The proof of Proposition \ref{prop_covariate} is left to Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@

Next, we describe our result for more than two tasks with same features, i.e. $X_i = X$ for any $i$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
\begin{theorem}[Many tasks]\label{thm_many_tasks}
%Suppose $X=Z\Sigma^{1/2}\in \R^{n\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho:=n/p>1$ being some fixed constant. Consider data models  $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $\e_i\in \R^{n}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. Moreover, assume that $X$, $\beta_i$ and $\e_i$ are all independent of each other.
	%Let $n = c \cdot p$.
	%Let $X\in\real^{n\times p}$ and $Y_i = X\beta_i + \varepsilon_i$, for $i = 1,\dots,k$.
%	Consider $t$ data models $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $X$ has covariance matrix $\Sigma$, and the entries of $\e_i$ are i.i.d. with mean zero and variance $\sigma^2$.
	%that satisfy Assumption \ref{assm_secA2} in the appendix.
	For the setting of $t$ tasks where $X_i = X$, for all $1\le i\le t$.
	Let $B^\star := [{\beta}_1,{\beta}_2,\dots,{\beta}_{t}]$ and $U_r\in\real^{t\times r}$ denote the linear model parameters.
	Let $U_r U_r^{\top}$ denote the best rank-$r$ subspace approximation of $(B^{\star})^\top\Sigma B^{\star}$.
	Assume that $\lambda_{\min}({B^{\star}}^\top\Sigma B^{\star})\gtrsim \sigma^2$.
	Let $v_i$ denote the $i$-th row vector of $U_r$.
	We have the following
	\squishlist
		\item  If	$\left(1 - \norm{v_t}^2 \right)\frac{\sigma^2}{\rho - 1} > \norm{\Sigma (B^{\star} U_r v_t - \beta_t)}^2 +  \oo \left( \|B^\star\|^2 + \sigma^2\right)$, then w.h.p $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$
		\item If $\left(1 - \norm{v_t}^2\right)\frac{\sigma^2}{\rho - 1} < \norm{\Sigma(B^{\star} U_r v_t - \beta_t)}^2 -  \oo \left( \|B^\star\|^2 + \sigma^2\right)$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{theorem}
Theorem \ref{thm_many_tasks} shows that when all tasks share the same features but have different labels, we can derive a sharp transition from positive to negative transfer.
Specially, $(1 - \norm{v_t}^2)\sigma^2/(\rho - 1)$ shows the amount of variance reduced using the multi-task estimator.
On the other hand, $\norm{\Sigma (B^{\star} U_r v_t - \beta_t)}$ shows the bias of the multi-task estimator.
The proof of \ref{thm_many_tasks} can be found in Appendix \ref{app_proof_many_tasks}.

>>>>

	With high probability, the variance of the multi-task estimator $\hat{\beta}_t^{\MTL}$ equals

@@

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Each task has sample size $n_i = \rho_i \cdot p$, for $i = 1, 2$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	We assume that for the target task, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma_i^2$, for $i=1,2$.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

In the isotropic model, we show that as we increase the distance between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL.
%We measure model dissimilarity as $\norm{\beta_1 - \beta_2}^2$, which is the distance between source and target in the isotropic model.
%Figure \ref{fig_model_shift} provides a simulation when $p = 200$.
%{The rest of parameter settings can be found in Appendix \ref{app_synthetic}.}
%Our result below will provide an explanation to this phenomenon.
%Based on Theorem \ref{thm_model_shift}, we derive the transition threshold in the following proposition.
We introduce the following notations.
{\small\begin{align*}
	\Psi(\beta_1, \beta_2) = {\ex{\bignorm{\beta_1 - \beta_2}^2}} / {\sigma^2},  \quad \Phi(\rho_1, \rho_2) = \frac{(\rho_1 + \rho_2 - 1)^2}{\rho_1 (\rho_1 + \rho_2) (\rho_2 - 1)}.
\end{align*}}

@@

\textit{Example.}
	We measure covariate shift by $M = \Sigma_1^{1/2} \Sigma_2^{-1/2}$.
	Assume that $\Psi(\beta_1, \beta_2) = 0$ for simplicity.
	We compare two cases: (i) when $M = \id$; (ii) when $M$ has $p/2$ singular values that are equal to $\lambda$ and $p/2$ singular values that are equal to $1 / \lambda$.
	Hence, $\lambda$ measures the severity of the covariate shift.
	Figure \ref{fig_covariate} shows a simulation of this setting by varying $\lambda$.
	We observe that as source/target sample ratio increases, the performance gap between the two cases increases.

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
We compare different choices of $M$ that belong to the following bounded set.
Let $\lambda_i$ be the $i$-th singular value of $M$.
Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.

{\small\begin{align*}
		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_i\le \mu_{\max}, \text{ for all } 1\le i\le p\right.},
\end{align*}}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.

\begin{proposition}[Covariate shift]\label{prop_covariate}
	Assume that $\Psi(\beta_1, \beta_2) = 0$ and $\rho_1, \rho_2>1$.
	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
	We have that
	{\small\[ g(\mu\id) \le \bigbrace{1+ \bigo{{\rho_2}/{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]}
\end{proposition}
This proposition shows that when source/target sample ratio is large, then having no covariate shift is optimal.
The proof of Proposition \ref{prop_covariate} is left to Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@

Next, we describe our result for more than two tasks with same features, i.e. $X_i = X$ for any $i$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
\begin{theorem}[Many tasks]\label{thm_many_tasks}
%Suppose $X=Z\Sigma^{1/2}\in \R^{n\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho:=n/p>1$ being some fixed constant. Consider data models  $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $\e_i\in \R^{n}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. Moreover, assume that $X$, $\beta_i$ and $\e_i$ are all independent of each other.
	%Let $n = c \cdot p$.
	%Let $X\in\real^{n\times p}$ and $Y_i = X\beta_i + \varepsilon_i$, for $i = 1,\dots,k$.
%	Consider $t$ data models $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $X$ has covariance matrix $\Sigma$, and the entries of $\e_i$ are i.i.d. with mean zero and variance $\sigma^2$.
	%that satisfy Assumption \ref{assm_secA2} in the appendix.
	For the setting of $t$ tasks where $X_i = X$, for all $1\le i\le t$.
	Let $B^\star := [{\beta}_1,{\beta}_2,\dots,{\beta}_{t}]$ and $U_r\in\real^{t\times r}$ denote the linear model parameters.
	Let $U_r U_r^{\top}$ denote the best rank-$r$ subspace approximation of $(B^{\star})^\top\Sigma B^{\star}$.
	Assume that $\lambda_{\min}({B^{\star}}^\top\Sigma B^{\star})\gtrsim \sigma^2$.
	Let $v_i$ denote the $i$-th row vector of $U_r$.
	There exists a value $\delta = \oo \left( \|B^\star\|^2 + \sigma^2\right)$ such that
	\squishlist
		\item  If	$\left(1 - \norm{v_t}^2 \right)\frac{\sigma^2}{\rho - 1} - \norm{\Sigma (B^{\star} U_r v_t - \beta_t)}^2 > \delta$, then w.h.p $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\left(1 - \norm{v_t}^2\right)\frac{\sigma^2}{\rho - 1} - \norm{\Sigma(B^{\star} U_r v_t - \beta_t)}^2 < -\delta$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{theorem}
Theorem \ref{thm_many_tasks} provides a sharp analysis of the bias-variance tradeoff beyond two tasks.
Specially, $(1 - \norm{v_t}^2)\sigma^2/(\rho - 1)$ shows the amount of reduced variance and $\norm{\Sigma (B^{\star} U_r v_t - \beta_t)}$ shows the bias of the multi-task estimator.
The proof of \ref{thm_many_tasks} can be found in Appendix \ref{app_proof_many_tasks}.
